Young's Experiment

### 2020-11-07 fdtd/em.tex
WIP 08Nov2020

In this article, we will consider electromagnetic waves in vacuum where there 
are no free charges or currents. Therefore $\rho_f = 0$ and $\vb{J}_f= 0$.
Maxwell equations, in this case, simplify to
\begin{eqnarray}
\div\vb{E} &=& 0 \label{s2e13} \\
\curl\vb{E} &=& -\pdv{\vb{B}}{t} \label{s2e14} \\
\div\vb{B} &=& 0 \label{s2e15} \\
\curl\vb{B} &=& \mu_0\epsilon_0\pdv{\vb{E}}{t} \label{s2e16}
\end{eqnarray}
Taking the curl of equation \eqref{s2e14} and using equation \eqref{s2e16}
in the result, we get
\begin{equation}\label{s2e17}
\laplacian\vb{E} = \mu_0\epsilon_0\pdv[2]{\vb{E}}{t}.
\end{equation}
This is the equation of an electromagnetic wave in vacuum. Its speed is
\begin{equation}\label{s2e18}
c_0 = \frac{1}{\sqrt{\epsilon_0\mu_0}}.
\end{equation}
If we take the curl of equation \eqref{s2e16} and use equation \eqref{s2e14},
we get
\begin{equation}\label{s2e19}
\laplacian\vb{B} = \mu_0\epsilon_0\pdv[2]{\vb{B}}{t}.
\end{equation}
Thus, the magnetic field $\vb{B}$ satisfies an identical equation. The key
physical idea governing the wave equations \eqref{s2e17} and \eqref{s2e19}
is that a changing magnetic field produces an electric field (Faraday's law
of equation \eqref{s2e14}) and a changing electric field produces a 
magnetic field (Ampere's law with Maxwell's correction of equation 
\eqref{s2e16}). These two laws together ensure that the electric and magnetic
waves sustain each other and produce an electromagnetic wave.

@@

We will illustrate FDTD for a simple one-dimensional problem. We observed
in section \ref{s2} that an electromagnetic wave is sustained because a 
changing electric field produces a magnetic field and vice versa. For sake
of simplicity, let us consider the progagation of electromagnetic waves in
vacuum. The pair of equations from which the wave equation emerges is
\begin{eqnarray}
\curl\vb{H} &=& \epsilon_0\pdv{\vb{E}}{t} \label{s5e3} \\
\curl\vb{E} &=& -\mu_0\pdv{\vb{H}}{t} \label{s5e4}
\end{eqnarray}
Let us assume that the wave propagates in the $z$-direction with the 
electric field in the $x$-direction and the $\vb{H}$ field in the
$y$-direction. Equations \eqref{s5e3} and \eqref{s5e4} simplify to
\begin{eqnarray}
\pdv{E_x}{t} &=& -\frac{1}{\epsilon_0}\pdv{H_y}{z} \label{s5e5} \\
\pdv{H_y}{t} &=& -\frac{1}{\mu_0}\pdv{E_x}{z} \label{s5e6}
\end{eqnarray}
The fields $E_x$ and $H_y$ are functions of $z$ and $t$. In this case, the
domain of the problem is a segment along the $z$-axis. We divide it in by a 
set of grid points and evaluate the fields at each one of them at successive
instants of time. The numerical solution of the problem thus consists in
evaluating $E_x(z, t)$ and $H_y(z, t)$ at a discrete set of points. We can
approximate the derivatives in equations \eqref{s5e5} and \eqref{s5e6} in
terms of central differences and write
\begin{eqnarray}
\frac{E_x(k, n + 1/2) - E_x(k, n - 1/2)}{\Delta t} 
 &=& -\frac{H_y(k + 1/2, n) - H_y(k - 1/2, n)}{\epsilon_0\Delta x}
\label{s5e7} \\
\frac{H_y(k + 1/2, n + 1) - H_y(k + 1/2, n)}{\Delta t} 
&=&
-\frac{E_x(k + 1, n + 1/2) - E_x(k - 1, n + 1/2)}{\epsilon_0\Delta x}
\nonumber \\
 & & \label{s5e8} 
\end{eqnarray}

Notes:
\begin{enumerate}
\item What is the Finite difference method?
\item What is FDTD? Why is it useful in solving electromagnetic problems?
\item Courant-Friedrichs-Lewy stability condition.
\item Illustration of an example in 1d.
\end{enumerate}

>>>>

In this article, we will consider electromagnetic waves in vacuum where there 
are no free charges or currents. Therefore $\rho_f = 0$ and $\vb{J}_f= 0$.
Maxwell equations, in this case, simplify to
\begin{eqnarray}
\div\vb{E} &=& 0 \label{s2e13} \\
\curl\vb{E} &=& -\pdv{\vb{B}}{t} \label{s2e14} \\
\div\vb{B} &=& 0 \label{s2e15} \\
\curl\vb{B} &=& \mu_0\epsilon_0\pdv{\vb{E}}{t} \label{s2e16}
\end{eqnarray}
Taking the curl of equation \eqref{s2e14} and using equation \eqref{s2e16}
in the result, we get
\begin{equation}\label{s2e17}
\laplacian\vb{E} = \mu_0\epsilon_0\pdv[2]{\vb{E}}{t}.
\end{equation}
This is the equation of an electromagnetic wave in vacuum. Its speed is
\begin{equation}\label{s2e18}
c = \frac{1}{\sqrt{\epsilon_0\mu_0}}.
\end{equation}
If we take the curl of equation \eqref{s2e16} and use equation \eqref{s2e14},
we get
\begin{equation}\label{s2e19}
\laplacian\vb{B} = \mu_0\epsilon_0\pdv[2]{\vb{B}}{t}.
\end{equation}
Thus, the magnetic field $\vb{B}$ satisfies an identical equation. The key
physical idea governing the wave equations \eqref{s2e17} and \eqref{s2e19}
is that a changing magnetic field produces an electric field (Faraday's law
of equation \eqref{s2e14}) and a changing electric field produces a 
magnetic field (Ampere's law with Maxwell's correction of equation 
\eqref{s2e16}). These two laws together ensure that the electric and magnetic
waves sustain each other and produce an electromagnetic wave.

@@

We will illustrate FDTD for a simple one-dimensional problem. We observed
in section \ref{s2} that an electromagnetic wave is sustained because a 
changing electric field produces a magnetic field and vice versa. For sake
of simplicity, let us consider the progagation of electromagnetic waves in
vacuum. The pair of equations from which the wave equation emerges is
\begin{eqnarray}
\curl\vb{H} &=& \epsilon_0\pdv{\vb{E}}{t} \label{s5e3} \\
\curl\vb{E} &=& -\mu_0\pdv{\vb{H}}{t} \label{s5e4}
\end{eqnarray}
Let us assume that the wave propagates in the $z$-direction with the 
electric field in the $x$-direction and the $\vb{H}$ field in the
$y$-direction. Equations \eqref{s5e3} and \eqref{s5e4} simplify to
\begin{eqnarray}
\pdv{E_x}{t} &=& -\frac{1}{\epsilon_0}\pdv{H_y}{z} \label{s5e5} \\
\pdv{H_y}{t} &=& -\frac{1}{\mu_0}\pdv{E_x}{z} \label{s5e6}
\end{eqnarray}
The fields $E_x$ and $H_y$ are functions of $z$ and $t$. In this case, the
domain of the problem is a segment along the $z$-axis. We divide it in by a 
set of grid points and evaluate the fields at each one of them at successive
instants of time. The numerical solution of the problem thus consists in
evaluating $E_x(z, t)$ and $H_y(z, t)$ at a discrete set of points. We can
approximate the derivatives in equations \eqref{s5e5} and \eqref{s5e6} in
terms of central differences and write \cite{sullivan2013electromagnetic}
\begin{eqnarray}
\frac{E_x(k, n + 1/2) - E_x(k, n - 1/2)}{\Delta t} 
 &=& -\frac{H_y(k + 1/2, n) - H_y(k - 1/2, n)}{\epsilon_0\Delta x}
\label{s5e7} \\
\frac{H_y(k + 1/2, n + 1) - H_y(k + 1/2, n)}{\Delta t} 
&=&
-\frac{E_x(k + 1, n + 1/2) - E_x(k - 1, n + 1/2)}{\epsilon_0\Delta x}
\nonumber \\
 & & \label{s5e8} 
\end{eqnarray}
We rewrite these equations as
\begin{eqnarray}
E_x\left(k ,n+\frac{1}{2}\right) &=& E_x\left(k ,
n-\frac{1}{2}\right) \nonumber \\
 & & -\frac{\Delta t}{\epsilon_0\Delta x}\left(\left(
H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
\right) \nonumber \\
 & & \label{s5e9}
\end{eqnarray}
and
\begin{eqnarray}
H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
\nonumber \\
 & & -\frac{\Delta t}{\mu_0\Delta x}\left(
E_x\left(k + 1, n + \frac{1}{2}\right) - E_x\left(k - 1, n+\frac{1}{2}\right)
\right) \nonumber \\
 & & \label{s5e10}
\end{eqnarray}
A few observations about these equations:
\begin{itemize}
\item $E_x$ is computed at space coordinates $0, 1, \ldots$ and time 
coordinates $1/2, 3/2, \ldots$.
\item $H_y$ is computed at space coordinates $1/2, 3/2, \ldots$ and time 
coordinates $0, 1, \ldots$.
\item The points at which $E_x$ is computed are interleaved between the points
at which $H_y$ is computed in both space and time grid.
\end{itemize}
When these equations are implemented in code, we compute the fields at 
\emph{all} points in the domain \emph{at a given instant of time}. We repeat
this process for the desired number of time steps.

There are a few more considerations when we implement these equations in code.
Equations \eqref{s5e9} and \eqref{s5e10} are very similar except that the
coefficient of the second term on the right hand side of the two equations
have $\epsilon_0$ in the first case and $\mu_0$ in the second case. The
values of these universal constants are $\epsilon_0 = 8.85418782\times 10^{-12}
$ m$^{-3}$kg$^{-1}$s$^4$A$^2$ and $\mu_0=1.25663706 \times 10^{-6}$ m kg
s$^{-2}$ A$^{-2}$. The two constants differ by $6$ orders of magnitude. In
order to bridge this gap, we introduce a new variable
\begin{equation}\label{s5e11}
\tilde{E}_x = \sqrt{\frac{\epsilon_0}{\mu_0}}E_x.
\end{equation}
In terms of this variable, equations \eqref{s5e9} and \eqref{s5e10} become
\begin{eqnarray}
\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \tilde{E}_x\left(k ,
n-\frac{1}{2}\right) \nonumber \\
 & & -\frac{\Delta t}{\mu_0\epsilon_0\Delta x}\left(\left(
H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
\right) \nonumber \\
 & & \label{s5e12}
\end{eqnarray}
and
\begin{eqnarray}
H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
\nonumber \\
 & & -\frac{\Delta t}{\mu_0\epsilon_0\Delta x}\left(
\tilde{E}_x\left(k + 1, n + \frac{1}{2}\right) - \tilde{E}_x\left(k - 1, n+\frac{1}{2}\right)
\right) \nonumber \\
 & & \label{s5e13}
\end{eqnarray}
Recall from equation \eqref{s2e18} that $c = 1/\sqrt{\epsilon_0\mu_0}$ so that
we can write the two equations above as
\begin{equation}\label{s5e14}
\tilde{E}_x = \sqrt{\frac{\epsilon_0}{\mu_0}}E_x.
\end{equation}
In terms of this variable, equations \eqref{s5e9} and \eqref{s5e10} become
\begin{eqnarray}
\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \tilde{E}_x\left(k ,
n-\frac{1}{2}\right) \nonumber \\
 & & -\frac{c \Delta t}{\Delta x}\left(\left(
H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
\right) \nonumber \\
 & & \label{s5e15}
\end{eqnarray}
and
\begin{eqnarray}
H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
\nonumber \\
 & & -\frac{c \Delta t}{\Delta x}\left(
\tilde{E}_x\left(k + 1, n + \frac{1}{2}\right) - \tilde{E}_x\left(k - 1, n+\frac{1}{2}\right)
\right) \nonumber \\
 & & \label{s5e16}
\end{eqnarray}
While implementing these equations in software we will have to choose certain
values for $\Delta x$ and $\Delta t$. $\Delta x$ is chosen in such a way that
there are at least ten points per wave. That is, if we are simulating a wave
with wavelength $\lambda$ then \cite{sullivan2013electromagnetic}
\begin{equation}\label{s5e17}
\Delta x \approx \frac{\lambda}{10}.
\end{equation}
The time step $\Delta t$ is determined by the Courant-Friedrichs-Lewy condition
\cite{cfl}. When we are solving a hyperbolic partial differential equation
numerically on a grid, we should choose a time step in such a way that it is
smaller than the time taken for the wave to travel to a neighboring point
a distance $\Delta x$ away. Electromagnetic waves need a time $\Delta x/c$
to travel from one grid point to the neighboring grid point. Therefore, we
choose 
\begin{equation}\label{s5e18}
\Delta t = \frac{\Delta x}{2c}.
\end{equation}
Using this equation in \eqref{s5e15} and \eqref{s5e16} we get
\begin{eqnarray}
\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \tilde{E}_x\left(k ,
n-\frac{1}{2}\right) \nonumber \\
 & & -\frac{1}{2}\left(\left(
H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
\right) \nonumber \\
 & & \label{s5e19}
\end{eqnarray}
and
\begin{eqnarray}
H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
\nonumber \\
 & & -\frac{1}{2}\left(
\tilde{E}_x\left(k + 1, n + \frac{1}{2}\right) - \tilde{E}_x\left(k - 1, n+\frac{1}{2}\right)
\right) \nonumber \\
 & & \label{s5e20}
\end{eqnarray}
Equations \eqref{s5e19} and \eqref{s5e20} can be readily written in software
to simulate propagation of electromagnetic waves in one dimension.

So far we have considered the propagation of electromagnetic waves in 
vacuum. We now consider their propagation in dielectric and lossy media. In
the case of a dielectric medium, the first equation of the pair \eqref{s5e5}
and \eqref{s5e6} changes to
\begin{equation}\label{s5e21}
\pdv{E_x}{t} = -\frac{1}{\epsilon}\pdv{H_y}{z}.
\end{equation}
For linear dielectric, one can write
\begin{equation}\label{s5e22}
\epsilon = \kappa\epsilon_0,
\end{equation}
where $\kappa > 1$ is the relative permittivity, also called the dielectric
constant. If we follow the development of equations \eqref{s5e19} and
\eqref{s5e20} from Maxwell equation with the changes indicated in equations
\eqref{s5e21} and \eqref{s5e22}, we get
\begin{eqnarray}
\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \tilde{E}_x\left(k ,
n-\frac{1}{2\kappa}\right) \nonumber \\
 & & -\frac{1}{2\kappa}\left(\left(
H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
\right) \nonumber \\
 & & \label{s5e23}
\end{eqnarray}
and
\begin{eqnarray}
H_y\left(k+\frac{1}{2},n+1\right) &=& H_y\left(k+\frac{1}{2}, n\right) 
\nonumber \\
 & & -\frac{1}{2}\left(
\tilde{E}_x\left(k + 1, n + \frac{1}{2}\right) - \tilde{E}_x\left(k - 1, n+\frac{1}{2}\right)
\right) \nonumber \\
 & & \label{s5e24}
\end{eqnarray}
Equation \eqref{s5e24} is identical with \eqref{s5e20} while equation
\eqref{s5e23} differs from \eqref{s5e21} by the presence of dielectric 
constant in the coefficient of the second term on the right hand side.

In the case of electromagnetic waves propagating in a medium of dielectric
constant $\kappa$ and electric conductivity $\sigma$, Ampere's law becomes
\begin{equation}\label{s5e25}
\curl\vb{H} = \kappa\epsilon_0\pdv{\vb{E}}{t} - \sigma\vb{E},
\end{equation}
where we have used Ohm's law \eqref{s4e5}. The one dimensional equations
for the $H_y$ and $\tilde{E}_x$ (refer to \eqref{s5e11}) fields are
\begin{eqnarray}
\pdv{\tilde{E}_x}{t} &=& -\frac{c}{\kappa}\pdv{H_y}{z} - 
\frac{\sigma}{\kappa\epsilon_0}\tilde{E}_x \label{s5e26} \\
\pdv{\tilde{H}_y}{t} &=& -c\pdv{\tilde{E}_x}{z}. \label{s5e27}
\end{eqnarray}
If we follow the derivation of the difference equation from the differential
equations and write 
\[
\tilde{E}_x(k, n) = \frac{\tilde{E}_x(k, n + 1/2) + \tilde{E}_x(k, n - 1/2)}{2}
\]
then we get a pair of equation of which the second one is the same as 
\eqref{s5e24} but the first one is
\begin{eqnarray}
\tilde{E}_x\left(k ,n+\frac{1}{2}\right) &=& \frac{1-\alpha}{1+\alpha}
\tilde{E}_x\left(k , n-\frac{1}{2\kappa}\right) \nonumber \\
 & & -\frac{1}{2\kappa(1 + \alpha)}\left(\left(
H_y\left(k + \frac{1}{2}, n\right) - H_y\left(k - \frac{1}{2}, n\right)\right)
\right), \nonumber \\
 & & \label{s5e28}
\end{eqnarray}
where \cite{sullivan2013electromagnetic}
\begin{equation}\label{s5e29}
\alpha = \frac{\sigma\Delta t}{2\kappa\epsilon_0}.
\end{equation}

\section{Some practical matters}\label{s6}
\subsection{Absorbing boundary condition}
\subsection{Perfectly matching layer}

\section{A simple example of FDTD}\label{s7}
